---
title: "Statistical analyses: Human winner and loser effects and rank-dependent optimal behaviour "
subtitle: "Noah Smith & Reuven Dukas"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(effectsize)
library(ggplot2);theme_set(theme_linedraw())
library(emmeans)
library(glmmTMB)
library(DHARMa)
library(ordinal)
```


\begin{center}
\large Switching Analysis
\end{center}

```{r}
WL_switching_df <- readRDS("WL_switching_df.rds")
summary(WL_switching_df)
```
\subsubsection{Information about switching dataframe}

  "switch.pref" levels correspond to: 
1: I strongly prefer to continue playing the shooting videogame.
2: I moderately prefer to continue playing the shooting videogame.
3: I have no preference for which videogame I play.
4: I moderately prefer to switch to playing the racing videogame.
5: I strongly prefer to switch to playing the racing videogame.

\subsubsection{Descriptive histogram of switching preferences for winners and losers}


```{r}
WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))

switch_labels <- c("1" = "stay (strong)", "2" = "stay (moderate)", 
                   "3" = "indifferent", "4" = "switch (moderate)", 
                   "5" = "switch (strong)")

strip_labels <- c("winner" = "Winner", "loser" = "Loser")

switch_barplot <- ggplot(WL_switching_df, aes(x = switch.pref,
                                              fill = switch.pref)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labels)) + 
  scale_x_discrete(labels = switch_labels) +
  labs(x = "Preference for shooter game (stay) or racing game (switch)",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 10),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(switch_barplot)

# ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/switching_barplot.png",
#       plot = switch_barplot, 
#       width = 12,
#       height = 8, 
#       dpi = 500)

WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))
```

\subsubsection{Cumulative link model of participant switching preference}

No participant ID as random effect because each participant provided only one response for 
switch.pref.

```{r}
switch_clm <- clm(switch.pref ~ 1 + treatment + age, link = "logit",
                  threshold = "flexible", Hess = TRUE, nAGQ = 10,
                  data = WL_switching_df)

null_switch_clm <- clm(switch.pref ~ 1, link = "logit",
                       threshold = "flexible", Hess = TRUE, nAGQ = 10,
                       data = WL_switching_df)

anova(switch_clm, null_switch_clm)
```
Fitted model with experiment and age has significantly better fit than null (intercept-only) model.

\subsubsection{Estimated marginal means plot of overall effect of treatment}

Emmeans plot and table of overall coefficient values, suggesting 
a statistically clear effect for greater switching preference in losers.

```{r}
switch_Ns <- c("n = 68", "n = 67")
emm_switch <- emmeans(switch_clm, ~treatment, type = "response")
plot(emm_switch) + theme_linedraw()  + 
  labs(x = "Switching preference (cumulative log odds)", y = "") +
  geom_text(label = switch_Ns, nudge_y = 0.2)
emm_switch
```
\subsubsection{Emmeans plot for model estimates of picking each level of switching}

```{r}
emm_switch_clm_allchoices <- emmeans(switch_clm, ~switch.pref|treatment, 
                                     mode = "prob")
plot(emm_switch_clm_allchoices) + 
  labs(x = "Probability of choice", 
       y = "Switching preference") + 
  scale_y_discrete(labels = switch_labels) +
  theme_linedraw()
```

Probability of picking each of the 5 levels after win vs. after loss. People generally
have a higher probability of wanting to switch when they're in the losing treatment 
(cumulative calculation is done above by emm_switch)

Bars are 95% confidence intervals.

\subsubsection{Summary table for main effects of treatment and age}

```{r}
summary(switch_clm)
```

Effect of treatment seems to have an influence on people's desire to switch, as 
shown in emmeans plot. Age has no effect.

\subsubsection{Ordinal effect size (cliff's delta) for switching preference}

We can also calculate an effect size measure (Cliff's delta) for the effect of 
winning vs losing on switching preference
```{r}
winner_switch_pref <- (WL_switching_df
                         %>% filter(treatment == "winner")
                         %>% pull(switch.pref))

loser_switch_pref <- (WL_switching_df
                         %>% filter(treatment == "loser")
                         %>% pull(switch.pref))

cliffs_delta(loser_switch_pref, winner_switch_pref)
```
Cliff's delta for switching = 0.31 (losers have "medium" sized greater preference to switch)

\newpage
\begin{center}
\large Waiting Analysis
\end{center}


```{r}
WL_waiting_df <- readRDS("WL_waiting_df.rds")
summary(WL_waiting_df)
```

\subsubsection{Information about waiting dataframe}

"wait.pref" levels correspond to:
1: I would like to start now.
2: I would like to wait 1 minute before playing.
3: I would like to wait 3 minutes before playing.
4: I would like to wait 5 minutes before playing.
5: I would like to wait 10 minutes before playing. 

"exp.level" corresponds to:
1: Participant has never played shooter videogames on a mouse and keyboard before.
2: Participant has played shooter videogames on a mouse and keyboard before.

\subsubsection{Descriptive histogram of waiting preferences:}
  
```{r}
WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))

wait_labels <- c("1" = "0", "2" = "1", "3" = "3", "4" = "5", "5" = "10")

wait_barplot <- ggplot(WL_waiting_df, aes(x = wait.pref, fill = wait.pref)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labels)) +
  scale_x_discrete(labels = wait_labels) +
  labs(x = "Waiting behaviour (minutes)",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 12),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(wait_barplot)

# ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/waiting_barplot.png",
#     plot = wait_barplot, 
#     width = 12,
#     height = 8, 
#     dpi = 500)

WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))
```

\subsubsection{Cumulative link model of waiting preferences using package "ordinal":}
  
```{r}
wait_clm <- clm(wait.pref ~ 1 + treatment + exp.level + age, link = "logit",
                threshold = "flexible", Hess = TRUE, nAGQ = 10,
                data = WL_waiting_df)

null_wait_clm <- clm(wait.pref ~ 1, link = "logit",
                     threshold = "flexible", Hess = TRUE, nAGQ = 10,
                     data = WL_waiting_df)

anova(wait_clm, null_wait_clm)
```

\subsubsection{Emmeans plot and table of main effect of winning and losing on waiting} 

```{r}
wait_Ns <- c("n = 69", "n = 69")
emm_wait_overall <- emmeans(wait_clm, ~treatment)

plot(emm_wait_overall) + theme_linedraw() + 
  labs(x = "Waiting behaviour (cumulative log odds)", y = "") +
  geom_text(label = wait_Ns, nudge_y = 0.2)
```
Suggests people in winning and losing treatments are waiting about equivalently.
Small and non-significant effect.

```{r}
emm_wait_clm_allchoices <- emmeans(wait_clm, ~wait.pref|treatment, mode = "prob")
plot(emm_wait_clm_allchoices) + theme_linedraw() + 
  labs(x = "Probability of choice", 
       y = "Waiting behaviour (higher number = longer requested waiting time)")
```

Probabilities for picking each of the wait time levels are quite similar for 
winners and losers. Bars are 95% confidence intervals.

\subsubsection{Summary table for coefficients and p-values of waiting CLM:}

```{r}
summary(wait_clm)
```

Effect of treatment does not have a statistically clear impact on 
waiting behaviour. Neither does age.

\subsubsection{Ordinal effect size (cliff's delta) for waiting behaviour}

We can also calculate Cliff's delta for the effect of losing vs. winning
on waiting behaviour.

```{r}
winner_wait_pref <- (WL_waiting_df
                         %>% filter(treatment == "winner")
                         %>% pull(wait.pref))

loser_wait_pref <- (WL_waiting_df
                         %>% filter(treatment == "loser")
                         %>% pull(wait.pref))

cliffs_delta(loser_wait_pref, winner_wait_pref)
```
Cliff's delta = 0.11 [95% CI overlaps 0] (losers have negligibly greater preference to wait)

\newpage
\begin{center}
\large Mood (PANAS) analysis
\end{center}

```{r}
WL_mood_long_df <- readRDS("WL_mood_long_df.rds")
summary(WL_mood_long_df)
```
\subsubsection{Mood long dataframe information:}

For each participant, we have 2 responses for positive affect 
(baseline and post-game) and 2 responses for negative affect (baseline and post-game). 
The raw mood dataframe / csv is the wide dataframe for mood.
In the long dataframe, "WL_mood_long_df", I have pivoted the values of these 
4 columns into a single column (affect.score). I moved "baseline" and "postgame"
to the column "timepoint", and I moved "positive affect" and "negative affect" 
to the column "affect type". The long dataframe therefore has 4 rows per participant.

\subsubsection{Descriptive plot of raw values for positive affect and negative affect}
```{r}
affect_labels <- c("positive" = "Positive affect", "negative" = "Negative affect", 
                    "positive" = "Positive affect", "negative" = "Negative affect")

raw_mood_boxplot <- ggplot(WL_mood_long_df, aes(x = affect.type, y = affect.score, fill = timepoint)) + 
                    geom_boxplot() + 
                    scale_x_discrete(labels = affect_labels) +
                    facet_wrap(~treatment, nrow = 1, labeller = as_labeller(strip_labels)) +
                    theme_linedraw() + 
                    labs(y = "Affect score",
                         x = "", fill = "Timepoint")

print(raw_mood_boxplot)
```

```{r}
WL_mood_percent_df <- readRDS("WL_mood_percent_df.rds")
summary(WL_mood_percent_df)
```
\subsubsection{Mood percent change from baseline dataframe info:} 

In the percent dataframe, I took the wide mood dataframe (raw csv) and converted
each participant's baseline and postgame negative affect and positive affect values
(4 columns total) into 2 "percent change from baseline" columns--1 for positive affect
and the other for negative affect. I then pivoted these columsn into long format, moving
the names to "affect.type" and the scores to "affect.percent.change". The percent
dataframe therefore has two rows per participants, one corresponding to the 
percent change in positive affect from baseline and the other corresponding to the 
percent change in negative affect from baseline. 

\subsubsection{Descriptive plot of percent change from baseline for positive affect and negative affect} 
```{r}
mood_percent_boxplot <- ggplot(WL_mood_percent_df, 
                        aes(x = affect.type, 
                            y = affect.percent.change,
                            fill = affect.type)) +
                        facet_wrap(~treatment, nrow = 1,
                                   labeller = as_labeller(strip_labels)) +
                        scale_x_discrete(labels = affect_labels) +
                        labs(y = "Percent change from baseline",
                             x = "Affect type") +
                        theme(legend.position = "none") +
                        geom_boxplot()

print(mood_percent_boxplot)
```
\subsubsection{Gaussian generalized linear mixed model (GLMM) of the mood percentage change data} 

We initially fit a GLMM to the long mood data that included a three-way interaction (treatment:affect.type:timepoint), but the fit was poor. We thus refit the model to the percent change dataframe and the overall fit was much better while
not changing our results. The GLMM includes participant ID as a random effect, which takes into account 
that each participant has a percent change response for positive affect and for negative affect. We
preferred the percent change model because the overall fit is much better and it changes little about 
our results. 

```{r}
aff_percent_change_glmm <- glmmTMB(affect.percent.change ~ treatment:affect.type + 
                                     experiment + age +
                                     (1|participant.ID),
                                    data = WL_mood_percent_df,
                                    family = "gaussian")

testResiduals(aff_percent_change_glmm)
testQuantiles(aff_percent_change_glmm)
```
\subsubsection{Estimated marginal means plots of model predictions} 

We can use the emmeans package to see what the model predicts. We specifically predicted
increase in positive affect for winners and increase in neg affect for losers. Points
are estimated marginal means based on model and bars are 95% confidence intervals.

```{r}
percent_change_emm <- emmeans(aff_percent_change_glmm, ~treatment:affect.type)
percent_change_emm

plot(percent_change_emm) + coord_flip() + geom_vline(xintercept = 0, lty = 3) +
  labs(y = "Treatment and affect type", x = "Percent change from baseline")

# just making this figure prettier

percent_change_emm_df <- as.data.frame(percent_change_emm)

strip_labels <- c("winner" = "Winner", "loser" = "Loser")
affect_labels <- c("positive" = "Positive affect", "negative" = "Negative affect", 
                    "positive" = "Positive affect", "negative" = "Negative affect")

percent_change_fig <- ggplot(percent_change_emm_df, aes(x = affect.type, y = emmean)) +
  geom_point(size = 3.5) +
  facet_wrap(~treatment, nrow=1, labeller = as_labeller(strip_labels)) + 
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0,
                linewidth = 1) +
  labs(x = "", y = "Percent change in affect from baseline") + 
  scale_x_discrete(labels = affect_labels) +
  geom_hline(yintercept = 0, lty = 3) +
  theme(axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 12),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(percent_change_fig)

# ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/percent_change_fig.png",
#       plot = percent_change_fig, 
#       width = 12,
#       height = 8, 
#       dpi = 500)
```

We can get z-values and p-values for these point estimates and confidence intervals using z-tests. Since the SEs and point estimates are from the GLMM, we can divide the point estimates by the SE to approximate a comparison to a null distribution (i.e. no change from baseline) based on the model.

```{r}
percent_change_emm_df_pvals <- (percent_change_emm_df 
                                %>% mutate(z.ratio = emmean/SE,
                                           p.value = 1.96*(1-pnorm(abs(z.ratio))))
)

percent_change_emm_df_pvals
```

\subsubsection{Main effects of experiment and age on percent change in affect}

We can also run the standard summary() function to get estimates for the main effects
of experiment and age on percent change in affect.
```{r}
summary(aff_percent_change_glmm)
```
\subsubsection{Effect size (Cohen's d) calculations for percentage change from baseline}

Here we are doing one-sample cohen's d calculations that compare the percentage
change values in our dataset (i.e. not modeled) to a null percentage change of 0.

```{r}
winner_pos_change <- (WL_mood_percent_df
                       %>% filter(treatment == "winner")
                       %>% filter(affect.type == "positive")
                       %>% pull(affect.percent.change))

cohens_d(winner_pos_change, mu = 0)

winner_neg_change <- (WL_mood_percent_df
                      %>% filter(treatment == "winner")
                      %>% filter(affect.type == "negative")
                      %>% pull(affect.percent.change))

cohens_d(winner_neg_change, mu = 0)

loser_pos_change <- (WL_mood_percent_df
                     %>% filter(treatment == "loser")
                     %>% filter(affect.type == "positive")
                     %>% pull(affect.percent.change))

cohens_d(loser_pos_change, mu = 0)

loser_neg_change <- (WL_mood_percent_df
                       %>% filter(treatment == "loser")
                       %>% filter(affect.type == "negative")
                       %>% pull(affect.percent.change))

cohens_d(loser_neg_change, mu = 0)
```