---
title: "Statistical analyses: Human winner and loser effects and rank-dependent optimal behaviour "
subtitle: "Noah Smith & Reuven Dukas"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(effectsize)
library(ggplot2);theme_set(theme_linedraw())
library(emmeans)
library(glmmTMB)
library(DHARMa)
library(ordinal)
```


\begin{center}
\large Switching Analysis
\end{center}


```{r}
WL_switching_df <- readRDS("WL_switching_df.rds")
summary(WL_switching_df)
```

about dataframe: 
  "switch.pref" levels correspond to: 
  1: I strongly prefer to continue playing the shooting videogame.
2: I moderately prefer to continue playing the shooting videogame.
3: I have no preference for which videogame I play.
4: I moderately prefer to switch to playing the racing videogame.
5: I strongly prefer to switch to playing the racing videogame.

Descriptive histogram of switching preferences for winners and losers. 

```{r}
WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))

switch_labels <- c("1" = "stay (strong)", "2" = "stay (moderate)", 
                   "3" = "indifferent", "4" = "switch (moderate)", 
                   "5" = "switch (strong)")

strip_labs <- c("winner" = "Winner", "loser" = "Loser")

switch_barplot <- ggplot(WL_switching_df, aes(x = switch.pref,
                                              fill = switch.pref)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labs)) + 
  scale_x_discrete(labels = switch_labels) +
  labs(x = "Preference for shooter game (stay) or racing game (switch)",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 10),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(switch_barplot)

# ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/switching_barplot.png",
#       plot = switch_barplot, 
#       width = 12,
#       height = 8, 
#       dpi = 500)

WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))
```


Cumulative link model using the package "ordinal". No participant 
ID as random effect because each participant provided only one response for 
switch.pref.

```{r}
switch_clm <- clm(switch.pref ~ 1 + treatment + age, link = "logit",
                  threshold = "flexible", Hess = TRUE, nAGQ = 10,
                  data = WL_switching_df)

null_switch_clm <- clm(switch.pref ~ 1, link = "logit",
                       threshold = "flexible", Hess = TRUE, nAGQ = 10,
                       data = WL_switching_df)

anova(switch_clm, null_switch_clm)
```

Emmeans plot and table of overall coefficient values, suggesting 
a statistically clear effect for greater switching preference in losers.

```{r}
switch_Ns <- c("n = 68", "n = 67")
emm_switch <- emmeans(switch_clm, ~treatment, type = "response")
plot(emm_switch) + theme_linedraw()  + 
  labs(x = "Switching preference (cumulative log odds)", y = "") +
  geom_text(label = switch_Ns, nudge_y = 0.2)
emm_switch
```


We can also calculate an effect size measure (Cliff's delta) for the effect of 
winning vs losing on switching preference
```{r}
# Cliff's delta for switching

winner_switch_pref <- (WL_switching_df
                         %>% filter(treatment == "winner")
                         %>% pull(switch.pref))

loser_switch_pref <- (WL_switching_df
                         %>% filter(treatment == "loser")
                         %>% pull(switch.pref))

cliffs_delta(loser_switch_pref, winner_switch_pref)

# cliff's delta for switching = 0.31 (losers have "medium" sized greater preference to switch)
```

Summary table for coefficient interpretation & p-values

```{r}
summary(switch_clm)
```

Effect of treatment seems to have an influence on people's desire to switch, 
where people have a stronger preference to switch to the other game after they've lost. 
Need to ask Ben more about how to interpret this. However, from what I understand, 
it's saying that, compared to the intercept ("win"), people in the losing treatment 
report a higher desire to switch (i.e. options "4" and "5").

```{r}
emm_switch_clm_allchoices <- emmeans(switch_clm, ~switch.pref|treatment, 
                                     mode = "prob")
plot(emm_switch_clm_allchoices) + 
  labs(x = "Probability of choice", 
       y = "Switching preference") + 
  scale_y_discrete(labels = switch_labels) +
  theme_linedraw()
```

Probability of picking each of the 5 levels after win vs. after loss. People generally
have a higher probability of wanting to switch when they're in the losing treatment 
(cumulative calculation is done above by emm_switch)

Bars are 95% confidence intervals.

\newpage
\begin{center}
\large Waiting Analysis
\end{center}


```{r}
WL_waiting_df <- readRDS("WL_waiting_df.rds")
summary(WL_waiting_df)
```

Information about the dataframe:
  "wait.pref" levels correspond to:
  1: I would like to start now.
2: I would like to wait 1 minute before playing.
3: I would like to wait 3 minutes before playing.
4: I would like to wait 5 minutes before playing.
5: I would like to wait 10 minutes before playing. 

"exp.level" corresponds to:
  1: I have never played shooter videogames on a mouse and keyboard before.
2: Between <1 and 5 hours of first person shooter videogame play on mouse and keyboard per week in the past 6 months
3: Greater than 5 hours of first person shooter videogame play on mouse and keyboard per week in the past 6 months

Descriptive histogram of waiting preferences:
  
```{r}
WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))

wait_labels <- c("1" = "0", "2" = "1", "3" = "3", "4" = "5", "5" = "10")

wait_barplot <- ggplot(WL_waiting_df, aes(x = wait.pref, fill = wait.pref)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labs)) +
  scale_x_discrete(labels = wait_labels) +
  labs(x = "Waiting behaviour (minutes)",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 12),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(wait_barplot)

# ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/waiting_barplot.png",
#     plot = wait_barplot, 
#     width = 12,
#     height = 8, 
#     dpi = 500)

WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))
```

Cumulative link model of waiting preferences using package "ordinal":
  
```{r}
wait_clm <- clm(wait.pref ~ 1 + treatment + exp.level + age, link = "logit",
                threshold = "flexible", Hess = TRUE, nAGQ = 10,
                data = WL_waiting_df)

null_wait_clm <- clm(wait.pref ~ 1, link = "logit",
                     threshold = "flexible", Hess = TRUE, nAGQ = 10,
                     data = WL_waiting_df)

anova(wait_clm, null_wait_clm)
```

Emmeans plot and table of waiting CLM, suggests people in winning 
and losing treatments are waiting about equivalently.
Small effect. If we had more data would be more statistically clear.

```{r}
wait_Ns <- c("n = 69", "n = 69")
emm_wait_overall <- emmeans(wait_clm, ~treatment)

plot(emm_wait_overall) + theme_linedraw() + 
  labs(x = "Waiting behaviour (cumulative log odds)", y = "") +
  geom_text(label = wait_Ns, nudge_y = 0.2)
```

We can also calculate Cliff's delta for the effect of losing vs. winning
on waiting behaviour.

```{r}
# Cliff's delta for effect of winning vs. losing on waiting

winner_wait_pref <- (WL_waiting_df
                         %>% filter(treatment == "winner")
                         %>% pull(wait.pref))

loser_wait_pref <- (WL_waiting_df
                         %>% filter(treatment == "loser")
                         %>% pull(wait.pref))

cliffs_delta(loser_wait_pref, winner_wait_pref)

# cliff's delta = 0.11 [95% CI overlaps 0] (losers have negligibly greater 
# preference to wait)
```

Summary table for coefficients and p-values of waiting CLM.
```{r}
summary(wait_clm)
```

Effect of treatment does not have a statistically clear impact on 
waiting behaviour.

```{r}
emm_wait_clm_allchoices <- emmeans(wait_clm, ~wait.pref|treatment, mode = "prob")
plot(emm_wait_clm_allchoices) + theme_linedraw() + 
  labs(x = "Probability of choice", 
       y = "Waiting behaviour (higher number = longer requested waiting time)")
```
Probabilities for picking each of the wait time levels are quite similar for 
winners and losers.

Bars are 95% confidence intervals.

\newpage
\begin{center}
\large Mood (PANAS) analysis
\end{center}


```{r}
WL_mood_long_df <- readRDS("WL_mood_long_df.rds")
summary(WL_mood_long_df)
```

mood df information: 

For each participant, We have 2 responses for positive affect 
(baseline and post-game) and 2 responses for negative affect (baseline and post-game). 
These are visible in the raw mood data, which is the wide dataframe for mood.
In the long dataframe, "WL_mood_long_df", I have pivoted the values of these 
4 columns into a single column (affect.score). I moved "baseline" and "postgame"
to the column "timepoint", and I moved "positive affect" and "negative affect" 
to the column "affect type". The long dataframe therefore has 4 rows per participant.

Descriptive plot of positive affect and negative affect

```{r}
WL_mood_long_df$treatment <- factor(WL_mood_long_df$treatment, 
                                    levels = rev(levels(WL_mood_long_df$treatment)))

WL_mood_long_df$affect.type <- factor(WL_mood_long_df$affect.type, 
                                    levels = rev(levels(WL_mood_long_df$affect.type)))

ggplot(WL_mood_long_df, aes(x = affect.type, y = affect.score, fill = timepoint)) + 
  geom_boxplot() + 
  facet_wrap(~treatment, nrow = 1, labeller = as_labeller(strip_labs)) +
  theme_linedraw() + 
  labs(y = "Affect score",
       x = "Affect type")

WL_mood_long_df$treatment <- factor(WL_mood_long_df$treatment, 
                                    levels = rev(levels(WL_mood_long_df$treatment)))

WL_mood_long_df$affect.type <- factor(WL_mood_long_df$affect.type, 
                                    levels = rev(levels(WL_mood_long_df$affect.type)))
```

Generalized linear mixed model assuming gaussian distribution using the package "glmmTMB" 

GLMM includes Participant ID as random effect, which takes into account the fact 
that each participant provides a positive affect response at baseline and postgame
and a negative affect response at baseline and postgame. Fit looks okay.

```{r}
aff_glmm <- glmmTMB(affect.score ~ treatment*timepoint*affect.type + 
                      experiment + age + (1|participant.ID),
                    data = WL_mood_long_df,family = "gaussian")

testResiduals(aff_glmm)
testQuantiles(aff_glmm)
```

Now look at emmeans plots to see what's going on. We specifically predicted
increase in positive affect for winners and increase in neg affect for losers.

```{r}
emm_aff_glmm <- emmeans(aff_glmm, ~treatment:timepoint:affect.type)
mood_Ns <- c("n = 137", "n = 136", "n = 137", "n = 136", "n = 137", "n = 136", "n = 137", "n = 136")
plot(emm_aff_glmm) + theme_linedraw() + labs(x = "Affect score", y = "") + 
  geom_text(label = mood_Ns, nudge_y = 0.3, size = 3.5)
```
Bars are 95% confidence intervals.

We can get p-values for baseline vs. post-game using pairwise contrasts. Because
of our three-way interaction term, this will generate a ton of p-values but we're only
interested in two, since our a priori predictions about mood were: 
1) wins would increase positive affect
2) losses would increase negative affect.

```{r}
emmean_aff_contrast <- emmeans(aff_glmm, pairwise ~ affect.type:treatment:timepoint)
emmean_aff_contrast
```

```{r}
# summary to check main effects of experiment and age

summary(aff_glmm)
```


Effect size (Cohen's d) calculations for effect of wins on pos and neg affect
and effect of losses on pos and neg affect:

```{r}
# Winner positive affect:

winner_pos_aff_baseline <- (WL_mood_long_df
                            %>% filter(treatment == "winner")
                            %>% filter(timepoint == "baseline")
                            %>% filter(affect.type == "positive")
                            %>% pull(affect.score))

winner_pos_aff_postgame <- (WL_mood_long_df
                            %>% filter(treatment == "winner")
                            %>% filter(timepoint == "postgame")
                            %>% filter(affect.type == "positive")
                            %>% pull(affect.score))

cohens_d(winner_pos_aff_postgame, winner_pos_aff_baseline)

# d = 0.47 (medium increase in pos affect after win)

# Winner negative affect:

winner_neg_aff_baseline <- (WL_mood_long_df
                            %>% filter(treatment == "winner")
                            %>% filter(timepoint == "baseline")
                            %>% filter(affect.type == "negative")
                            %>% pull(affect.score))

winner_neg_aff_postgame <- (WL_mood_long_df
                            %>% filter(treatment == "winner")
                            %>% filter(timepoint == "postgame")
                            %>% filter(affect.type == "negative")
                            %>% pull(affect.score))

cohens_d(winner_neg_aff_postgame, winner_neg_aff_baseline)

# d = -0.11, negligible decrease in neg affect after a win.

# Loser positive affect:

loser_pos_aff_baseline <- (WL_mood_long_df
                            %>% filter(treatment == "loser")
                            %>% filter(timepoint == "baseline")
                            %>% filter(affect.type == "positive")
                            %>% pull(affect.score))

loser_pos_aff_postgame <- (WL_mood_long_df
                            %>% filter(treatment == "loser")
                            %>% filter(timepoint == "postgame")
                            %>% filter(affect.type == "positive")
                            %>% pull(affect.score))

cohens_d(loser_pos_aff_postgame, loser_pos_aff_baseline)

# d = -0.11, negligible decrease in pos affect after a loss. 

# Loser negative affect:

loser_neg_aff_baseline <- (WL_mood_long_df
                           %>% filter(treatment == "loser")
                           %>% filter(timepoint == "baseline")
                           %>% filter(affect.type == "negative")
                           %>% pull(affect.score))

loser_neg_aff_postgame <- (WL_mood_long_df
                           %>% filter(treatment == "loser")
                           %>% filter(timepoint == "postgame")
                           %>% filter(affect.type == "negative")
                           %>% pull(affect.score))

cohens_d(loser_neg_aff_postgame, loser_neg_aff_baseline)

# d = 0.48, medium increase in neg affect after a loss.
```
Making dataframe for inferential fig on percent change in positive and negative 
affect from baseline based on mood GLMM

```{r}
emm_aff_glmm_df <- as.data.frame(emm_aff_glmm)
emm_aff_glmm_df

# All values used for the calculations below come from the above dataframes, which 
# is from the emmeans objects generated from the mood glmm

# Formula for calculating percent change in affect from baseline: 

# ((postgame - baseline)/baseline)*100

# Formula for calculating 95% confidence interval for percent change in affect from baseline:

# ((1.96*SE)/((baseline + post)/2))*100

# ------------------CHANGE IN POSITIVE AFFECT (WINNERS)--------------------

# percent change in positive affect for winners: 

(((32.45769-28.67092)/28.67092)*100) # 13.2077% change in pos affect for winners

# Confidence interval for change in positive affect. The CI is calculated
# using the wald method (hence the 1.96*SE) and averaging the mean estimates 
# across baseline and postgame (i.e., the confidence intervals are an 
# approximation of the change). 

# Formula: 95% CI = ((1.96*SE)/((baseline + post)/2))*100

((1.96*0.5297711)/((28.67092+32.45769)/2))*100 # ± 3.397268% CI

# # ------------------CHANGE IN POSITIVE AFFECT (LOSERS)--------------------

(((27.49314-28.23767)/28.23767)*100) # -2.636655% change in pos affect for losers
# CI for change in positive affect for losers
((1.96*0.5278298)/((28.23767+27.49314)/2))*100 # ± 3.712655 CI

# ------------------CHANGE IN NEGATIVE AFFECT (WINNERS)--------------------


# percent change in negative affect for winners: 
(((14.20768-14.66357)/14.66357)*100) # -3.108997% change in neg affect for winners
# CI for change in negative affect for winners
((1.96*0.5297711)/((14.20768+14.66357)/2))*100 # ± 7.192978% CI

# ------------------CHANGE IN NEGATIVE AFFECT (LOSERS)--------------------

# percent change in negative affect for losers: 
(((17.71942-15.15007)/15.15008)*100) # 16.95932% change in neg affect for losers
# CI for change in negative affect for losers
((1.96*0.5278298)/((15.15008+17.71942)/2))*100 # ± 6.294872% CI 

# Making dataframe to construct figure, all values come from the above
# calculations

affect_change_df <- tibble(
  affect = c("Positive affect","Positive affect","Negative affect","Negative affect"),
  treatment = c("Winner","Loser","Winner","Loser"),
  mean_change = c(13.2077,-2.636655,-3.108997,16.95933),
  CI_upper = c(13.2077+3.397268,-2.636655+3.712655,-3.108997+7.192978,16.95932+6.294872),
  CI_lower = c(13.2077-3.397268,-2.636655-3.712655,-3.108997-7.192978,16.95932-6.294872)
)

affect_change_df <- (affect_change_df
                     %>% mutate(treatment = fct_rev(treatment))
                     %>% mutate(affect = fct_rev(affect))
)

affect_change_fig <- ggplot(affect_change_df, aes(x = affect, y = mean_change)) +
  geom_point(size = 3.5) +
  facet_wrap(~treatment, nrow=1) + 
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                width = 0,
                linewidth = 1) +
  labs(x = "", y = "Percent change in affect from baseline") + 
        geom_hline(yintercept = 0, lty = 3) +
        theme(axis.title.x = element_text(size = 18, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 18),
        axis.text.x = element_text(size = 12),   
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 18))

print(affect_change_fig)

#ggsave(filename = "/Users/noahsmith/WL_RDOB/figures/affect_change_fig.png",
#       plot = affect_change_fig, 
#       width = 12,
#       height = 8, 
#       dpi = 500)

```
