---
title: "Statistical analyses: Adaptive significance of winner and loser effects: rank-dependent optimal behaviour"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(DHARMa)
library(effectsize)
library(emmeans)
library(ggplot2);theme_set(theme_linedraw())
library(glmmTMB)
library(lme4)
library(ordinal)
library(tidyverse)
```

\begin{center}
\large Switching Analysis
\end{center}

```{r}
# Load and inspect switching dataframe

WL_switching_df <- readRDS("WL_switching_df.rds")
summary(WL_switching_df)
```
\subsubsection{Information about switching dataframe}

"switch.pref" levels correspond to: 
1: I strongly prefer to continue playing the shooting videogame.
2: I moderately prefer to continue playing the shooting videogame.
3: I have no preference for which videogame I play.
4: I moderately prefer to switch to playing the racing videogame.
5: I strongly prefer to switch to playing the racing videogame.

\subsubsection{Descriptive histogram of switching preferences for winners and losers}

```{r}
WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))

switch_labels <- c("1" = "stay (strong)", "2" = "stay (moderate)", 
                   "3" = "indifferent", "4" = "switch (moderate)", 
                   "5" = "switch (strong)")

strip_labels <- c("winner" = "Winner", "loser" = "Loser")

switch_barplot <- ggplot(WL_switching_df, aes(x = switch.pref,
                                              fill = switch.pref)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labels)) + 
  scale_x_discrete(labels = switch_labels) +
  scale_fill_manual(values = c("5" = "#045A8D", "4" = "#2B8CBE","3" = "grey", "2" = "#D6604D","1" = "#B2182B")) +
  labs(x = "Preference for shooter game (stay) or racing game (switch)",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 20, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 20, margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 10),   
        axis.text.y = element_text(size = 16),
        strip.text = element_text(size = 24))

print(switch_barplot)

WL_switching_df$treatment <- factor(WL_switching_df$treatment, 
                                    levels = rev(levels(WL_switching_df$treatment)))
```

\subsubsection{Cumulative link model of participant switching preference}

```{r}
switch_clm <- clm(switch.pref ~ 1 + treatment + age, link = "logit",
                  threshold = "flexible", Hess = TRUE, nAGQ = 10,
                  data = WL_switching_df)

# creating null model and performing ANOVA between fitted and null model to see 
# whether fit is improved in our fitted model (using AIC)

null_switch_clm <- clm(switch.pref ~ 1, link = "logit",
                       threshold = "flexible", Hess = TRUE, nAGQ = 10,
                       data = WL_switching_df)

anova(switch_clm, null_switch_clm)

# AIC of fitted model is significantly lower, indicating fit of full model is 
# better than fit of null, intercept only model.

# then we can use summary function to see whether our predictors significantly affect
# our response variable.

summary(switch_clm)
```
\subsubsection{Estimated marginal means plot of overall effect of treatment}

Emmeans table and plot of treatment coefficient values for switching pref

```{r}
emm_switch <- emmeans(switch_clm, ~treatment, type = "response")
emm_switch_df <- as.data.frame(emm_switch)
emm_switch_df$treatment <- fct_rev(emm_switch_df$treatment)

emm_switch_df
trt_colors <- c(Loser = "blue", Winner = "red")

logodds_switching_fig <- ggplot(emm_switch_df, aes(y = emmean, x = treatment)) +
  geom_point(size = 5, color = trt_colors) +
  geom_errorbar(aes(ymin = asymp.LCL,
                    ymax = asymp.UCL),
                color = trt_colors,
                width = 0, linewidth = 1.25) + 
  labs(x = "", y = "Cumulative log odds of switching preference") + 
  scale_x_discrete(labels = strip_labels) +
  theme(axis.title.y = element_text(size = 24, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 24),   
        axis.text.y = element_text(size = 18),
        strip.text = element_text(size = 22))

print(logodds_switching_fig)
```
\subsubsection{Emmeans plot of switching model: log odds estimates for picking each level of switching}

```{r}
emm_switch_clm_allchoices <- emmeans(switch_clm, ~switch.pref|treatment, 
                                     mode = "prob")
plot(emm_switch_clm_allchoices) + 
  labs(x = "Probability of choice", 
       y = "Switching preference") + 
  scale_y_discrete(labels = switch_labels) +
  theme_linedraw()
```

Probability of picking each of the 5 levels after win vs. after loss. People generally
have a higher probability of wanting to switch when they're in the losing treatment 
(cumulative calculation is done above by emm_switch). Bars are 95% confidence intervals.
Can see that probability of "switch (strong)" and "switch (moderate)" are greater 
in the losing treatment.

Effect of treatment seems to have an influence on people's desire to switch, as 
shown in emmeans plot.

\subsubsection{Ordinal effect size (cliff's delta) for switching preference}

Cliff's delta calculation for switching preference based on treatment 

```{r}
# pulling winner switching preference and loser switching preference.

winner_switch_pref <- (WL_switching_df
                       %>% filter(treatment == "winner")
                       %>% pull(switch.pref))
  
loser_switch_pref <- (WL_switching_df
                      %>% filter(treatment == "loser")
                      %>% pull(switch.pref))

# calculating cliff's delta between loser and winner switching preference. 

cliffs_delta(loser_switch_pref, winner_switch_pref)
```

Cliff's delta for switching = 0.31

\newpage
\begin{center}
\large Waiting Analysis
\end{center}

```{r}
# Load and inspect waiting dataframe

WL_waiting_df <- readRDS("WL_waiting_df.rds")
summary(WL_waiting_df)
```

\subsubsection{Information about waiting dataframe}

"wait.pref" levels correspond to:
0: I would like to start now and play against the same opponent.
1: I would like to wait and play against a different opponent.

"exp.level" corresponds to:
1: Participant has never played shooter videogames on a mouse and keyboard before.
2: Participant has played shooter videogames on a mouse and keyboard before.

\subsubsection{Descriptive histogram of waiting preferences:}
  
```{r}
WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))

wait_labels <- c("start" = "Start", "wait" = "Wait")

wait_barplot <- ggplot(WL_waiting_df, aes(x = wait.spell, fill = wait.spell)) + 
  geom_bar() + facet_wrap(~treatment, nrow=1, 
                          labeller = as_labeller(strip_labels)) +
  scale_x_discrete(labels = wait_labels) +
  scale_fill_manual(values = c("wait" = "#045A8D", "start" = "#B2182B")) +
  labs(x = "Waiting behaviour",
       y = "Count") + 
  theme(legend.position = "none",
        axis.title.x = element_text(size = 20, 
                                    margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size = 20, margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 18),   
        axis.text.y = element_text(size = 18),
        strip.text = element_text(size = 24))


print(wait_barplot)

WL_waiting_df$treatment <- factor(WL_waiting_df$treatment, 
                                  levels = rev(levels(WL_waiting_df$treatment)))

```

\subsubsection{binomial GLM for waiting data:}
  
```{r}
waiting_glm <- glm(wait.pref ~ treatment + age + exp.level, family = "binomial", data = WL_waiting_df)

plot(simulateResiduals(waiting_glm))
testDispersion(waiting_glm)

# fit of model looks very good.

# we can also use the summary function to see the main effects of the model.

summary(waiting_glm)
```
\subsubsection{Effect sizes (log odds) for waiting choice of winners vs. losers}

```{r}
emm_waiting <- emmeans(waiting_glm, ~treatment)

# losers' log odds of waiting was 0.68, winners' log odds of waiting was -2.04

emm_waiting_df <- as.data.frame(emm_waiting)
emm_waiting_df

# just flipping treatment columns so that winner comes first in the figure.

emm_waiting_df$treatment <- fct_rev(emm_waiting_df$treatment)

logodds_waiting_fig <- ggplot(emm_waiting_df, aes(y = emmean, x = treatment)) +
  geom_point(size = 5, color = trt_colors) + geom_errorbar(aes(ymin = asymp.LCL,
                    ymax = asymp.UCL),
                color = trt_colors,
                width = 0, linewidth = 1.25) + 
  labs(x = "", y = "Log odds of waiting") + 
  scale_x_discrete(labels = strip_labels) +
  theme(axis.title.y = element_text(size = 24, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 24),   
        axis.text.y = element_text(size = 18),
        strip.text = element_text(size = 22))

plot(logodds_waiting_fig)
```
\newpage
\begin{center}
\large Mood (PANAS) analysis
\end{center}

```{r}
# load and inspect mood dataframe

WL_mood_long_df <- readRDS("WL_mood_long_df.rds")
summary(WL_mood_long_df)
```

\subsubsection{Mood long dataframe information:}

For each participant, we have 2 responses for positive affect 
(baseline and post-game) and 2 responses for negative affect (baseline and post-game). 
The raw mood dataframe / csv is the wide dataframe for mood.
In the long dataframe, "WL_mood_long_df", I have pivoted the values of these 
4 columns into a single column (affect.score). I moved "baseline" and "postgame"
to the column "timepoint", and I moved "positive affect" and "negative affect" 
to the column "affect type". The long dataframe therefore has 4 rows per participant.

\subsubsection{Descriptive plot of raw values for positive affect and negative affect}
```{r}
affect_labels <- c("positive" = "Positive affect", "negative" = "Negative affect", 
                   "positive" = "Positive affect", "negative" = "Negative affect")

raw_mood_boxplot <- ggplot(WL_mood_long_df, aes(x = affect.type, y = affect.score, fill = timepoint)) + 
  geom_boxplot() + 
  scale_x_discrete(labels = affect_labels) +
  facet_wrap(~treatment, nrow = 1, labeller = as_labeller(strip_labels)) +
  theme_linedraw() + 
  labs(y = "Affect score",
       x = "", fill = "Timepoint") + 
  theme(axis.title.y = element_text(size = 26, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 18),   
        axis.text.y = element_text(size = 24),
        strip.text = element_text(size = 26),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 24))

print(raw_mood_boxplot)

```
To conduct the analysis on percent change of mood from baseline, I will use the mood percent change
dataframe that I constructed in the data_cleaning.r file. 

```{r}
WL_mood_percent_df <- readRDS("WL_mood_percent_df.rds")
summary(WL_mood_percent_df)
```
\subsubsection{Mood percent change from baseline dataframe info:} 

In the percent dataframe, I took the wide mood dataframe (raw csv) and converted
each participant's baseline and postgame negative affect and positive affect values
(4 columns total) into 2 "percent change from baseline" columns--1 for positive affect
and the other for negative affect. I then pivoted these columns into long format, moving
the names to "affect.type" and the scores to "affect.percent.change". The percent
dataframe therefore has two rows per participants, one corresponding to the 
percent change in positive affect from baseline and the other corresponding to the 
percent change in negative affect from baseline. 

\subsubsection{Descriptive plot of percent change from baseline for positive affect and negative affect} 
```{r}
percent_mood_boxplot <- ggplot(WL_mood_percent_df, 
                        aes(x = affect.type, 
                            y = affect.percent.change,
                            fill = affect.type)) +
                        facet_wrap(~treatment, nrow = 1,
                                   labeller = as_labeller(strip_labels)) +
                        scale_x_discrete(labels = affect_labels) +
                        labs(y = "Percent change from baseline",
                             x = "") +
                        theme(legend.position = "none") +
                        geom_boxplot() + 
  theme(axis.title.y = element_text(size = 26, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 22),   
        axis.text.y = element_text(size = 24),
        strip.text = element_text(size = 26))

print(percent_mood_boxplot)
```
\subsubsection{Gaussian generalized linear mixed model (GLMM) of the mood percentage change data} 

```{r}
aff_percent_change_glmm <- glmmTMB(affect.percent.change ~ treatment:affect.type + 
                                     experiment + age + (1|participant.ID),
                                    data = WL_mood_percent_df,
                                    family = "gaussian")

testResiduals(aff_percent_change_glmm)
testQuantiles(aff_percent_change_glmm)

# fit looks fine, nothing to be too concerned about. Use summary() function
# to seem main effects of experiment and age.

summary(aff_percent_change_glmm)

# Detect no effect of experiment or age on how affect changes from baseline, 
# which makes sense given how similar the experiments are and how tight the age
# distribution is.

# The summary function is difficult to interpret for our specific predictions about 
# how positive affect and negative affect change based on win or loss. 
# So it's better that we just construct emmeans plots of percentage change from 
# baseline based on each of the treatments.
```
\subsubsection{Estimated marginal means plots of model predictions} 

We can use the emmeans package to see what the model predicts about mood change and how this 
aligns with our own a priori predictions. We specifically predicted
increase in positive affect for winners and increase in neg affect for losers. Points
are estimated marginal means based on model and bars are 95% confidence intervals.

```{r}
percent_change_emm <- emmeans(aff_percent_change_glmm, ~treatment:affect.type)
percent_change_emm

percent_change_emm_df <- as.data.frame(percent_change_emm)

strip_labels <- c("winner" = "Winner", "loser" = "Loser")
affect_labels <- c("positive" = "Positive affect", "negative" = "Negative affect", 
                    "positive" = "Positive affect", "negative" = "Negative affect")
mood_trt_colors <- c(winner = "red", loser = "blue", winner = "red", loser = "blue")
  
percent_change_fig <- ggplot(percent_change_emm_df, aes(x = affect.type, y = emmean)) +
  geom_point(size = 5, color = mood_trt_colors) +
  facet_wrap(~treatment, nrow=1, labeller = as_labeller(strip_labels)) + 
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                color = mood_trt_colors,
                width = 0,
                linewidth = 1.25) +
  labs(x = "", y = "Change from baseline (%)") + 
  scale_x_discrete(labels = affect_labels) +
  geom_hline(yintercept = 0, lty = 3) +
  theme(axis.title.y = element_text(size = 26, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        axis.text.x = element_text(size = 22),   
        axis.text.y = element_text(size = 24),
        strip.placement = "outside",
        strip.text = element_text(size = 26))

print(percent_change_fig)
```

We can get z-values and p-values for these point estimates and confidence intervals using z-tests. Since the SEs and point estimates are from the GLMM, we can divide the point estimates by the SE to calculate a comparison to a null distribution (i.e. no change from baseline) based on the model.

```{r}
percent_change_emm_df_pvals <- (percent_change_emm_df 
                                %>% mutate(z.ratio = emmean/SE,
                                           p.value = 1.96*(1-pnorm(abs(z.ratio))))
)

percent_change_emm_df_pvals
```

\subsubsection{Effect size (Cohen's d) calculations for baseline to postgame of pos and neg affect for winners and losers}

```{r}
# Winner positive affect:
winner_pos_aff_baseline <- (WL_mood_long_df
  %>% filter(treatment == "winner")
  %>% filter(timepoint == "baseline") 
  %>% filter(affect.type == "positive") 
  %>% pull(affect.score))

winner_pos_aff_postgame <- (WL_mood_long_df
  %>% filter(treatment == "winner")
  %>% filter(timepoint == "postgame") 
  %>% filter(affect.type == "positive") 
  %>% pull(affect.score))

cohens_d(winner_pos_aff_postgame, winner_pos_aff_baseline)

# Winner negative affect:
winner_neg_aff_baseline <- (WL_mood_long_df
  %>% filter(treatment == "winner")
  %>% filter(timepoint == "baseline") 
  %>% filter(affect.type == "negative") 
  %>% pull(affect.score))

winner_neg_aff_postgame <- (WL_mood_long_df
  %>% filter(treatment == "winner")
  %>% filter(timepoint == "postgame") 
  %>% filter(affect.type == "negative")
  %>% pull(affect.score))

cohens_d(winner_neg_aff_postgame, winner_neg_aff_baseline)

# Loser positive affect:

loser_pos_aff_baseline <- (WL_mood_long_df
  %>% filter(treatment == "loser")
  %>% filter(timepoint == "baseline") 
  %>% filter(affect.type == "positive") 
  %>% pull(affect.score))

loser_pos_aff_postgame <- (WL_mood_long_df
  %>% filter(treatment == "loser")
  %>% filter(timepoint == "postgame") 
  %>% filter(affect.type == "positive") 
  %>% pull(affect.score))

cohens_d(loser_pos_aff_postgame, loser_pos_aff_baseline)

# Loser negative affect:

loser_neg_aff_baseline <- (WL_mood_long_df
  %>% filter(treatment == "loser")
  %>% filter(timepoint == "baseline") 
  %>% filter(affect.type == "negative") 
  %>% pull(affect.score))

loser_neg_aff_postgame <- (WL_mood_long_df
  %>% filter(treatment == "loser")
  %>% filter(timepoint == "postgame") 
  %>% filter(affect.type == "negative") 
  %>% pull(affect.score))

cohens_d(loser_neg_aff_postgame, loser_neg_aff_baseline)
```